---
title: "RegioMSI v1.1 tutorial (Updated 20241210)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# RegioMSI v1.1: Parallel processing and analysis of mass spectrometry imaging-based metabolomics datasets

### **Important:**
#### - Set **"eval ="** to **"F"** to skip execution of code chunk.
#### - Recommended hardware specs: Windows >10 with 128Gb of RAM OR Linux Ubuntu >22.04 with 128Gb of RAM
#### - Use parallel::mclapply and set core number to run compatible functions in parallel (not covered by this tutorial but examples can be found in package documentation)

## 1.1. Installation of package and dependencies
```{r inst, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=F}
# Install devtools and BiocManager
install.packages("devtools")
install.packages("BiocManager")

# Install RegioMSI
## NOTE: Bioconductor libraries are not installed by default; see package
## documentation for further details on Bioconductor packages required by
## RegioMSI
devtools::install_github(
  "cschasestevens/RegioMSI",
  ref = "master",
  build_vignettes = TRUE
)

# Install Bioconductor packages with the following:
BiocManager::install("Cardinal")

## Ensure that working directory is set to project folder with setwd()
## The project folder should contain the following subfolders:
## analysis, processed, ref, and data
```

## 2.1. Data Import
#### Loads all .imzML files from specified folder 
```{r impt, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
library(RegioMSI)
# Dataset input parameters (replace with local file location)
# Use msi_load_data() to load all samples present in the chosen directory
lp1 <- msi_input_param(
  dp = "data_test/package/",
  pol = "pos",
  mz_range = c(300, 1300),
  mz_res = 5
)
# Load all .imzML
data("testd", package = "RegioMSI")
data("an1", package = "RegioMSI")
data("md1", package = "RegioMSI")
d <- testd
```

## 2.2. Processing: Peak Detection, Alignment, Binning, and Annotation
#### Performs peak detection, alignment, binning, peak annotation, and assignment for all study samples. Detection and alignment are performed simultaneously to produce a single peak list for the entire dataset. Peaks from each sample are binned to the resulting reference list to convert data from profile to centroid and are used for downstream analyses. Peak annotations are optional but generally correspond to a list of compounds validated by LC-MS/MS.
```{r proc, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
#---- Peak detection ----
# Perform all peak processing;
# Previous versions of Cardinal required
# a separate peak binning step, which is
# no longer applicable in newer versions
d2 <- msi_peak_proc(
  ld = d[["Data.files"]],
  lp = lp1,
  parl = FALSE,
  core_perc = 0.2
)
d2
names(d2)

#---- Peak annotation ----
# Requires:
# - Processed peaks from previous step
# - annotation text file for assigning annotations to peaks
## Annotate and filter known peaks
d2a <- msi_peak_anno(
  ldp = d2[["Processed.Peaks"]],
  la = an1
)
names(d2a)
d2a[["Data.filtered"]]
d2a[["Data.original"]]

# Check data
Cardinal::imageData(d2a[["Data.filtered"]])
Cardinal::features(d2a[["Data.filtered"]])
Cardinal::pixelData(d2a[["Data.filtered"]])
Cardinal::run(d2a[["Data.filtered"]])
Cardinal::spectra(
  d2a[["Data.filtered"]],
  "intensity"
)[, 1:25]
d2a[["Data.filtered"]]
```

## 2.3. Post-processing: Artifact Removal and Normalization
#### Converts MSI experiment objects produced by Cardinal R to dense matrices after annotating peaks. This increases memory consumption but enables greater compatibility and faster calculation of normalization methods from other R packages. Two different normalization methods are tested in this script and compared to pre-normalized data, traditional total ion current (TIC) normalization and sparse locally estimated scatterplot smoothing (sLOESS) normalization. sLOESS is adapted from LOESS normalization in untargeted LC-MS/MS-based metabolomics assays and performs LOESS normalization for every non-zero intensity pixel in each image across multiple sample runs. This method addresses intra- and intersample signal drift in MSI experiments whereas TIC normalization in MSI is traditionally implemented on a per-sample basis and does not directly account for systematic variation in signal intensity.
```{r norm, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
#---- Format MSImagingExperiment ----
# Load processed data, annotation list, and metadata table (see example)
names(d2a)
Cardinal::imageData(d2a[["Data.original"]])
## Format
d_norm <- msi_norm_form(
  ldp = d2a[["Data.original"]],
  la = d2a[["All"]],
  md = md1,
  ftype = "all"
)
names(d_norm)
head(d_norm[["Data"]])
head(d_norm[["Feature"]])
head(d_norm[["Pixel"]])

#---- Artifact Removal ----
# Calculate descriptive stats
data_qc <- msi_data_qc(
  dm = d_norm[["Data"]],
  feat = d_norm[["Feature"]],
  pix = d_norm[["Pixel"]],
  parl = FALSE,
  core_perc = 0.5
)
head(data_qc)

# Remove flagged artifacts from downstream normalization
d_norm[["Data"]] <- as.matrix(
  as.data.frame(d_norm[["Data"]])[, data_qc[["artifact.flag"]] == FALSE]
)
ncol(d_norm[["Data"]])
d_norm[["Feature"]] <- d_norm[["Feature"]][
  data_qc[["artifact.flag"]] == FALSE,
]
nrow(d_norm[["Feature"]])
#---- Perform Normalization ----
names(d_norm)
# For datasets with ~ 100 compounds or more, it is often necessary to
# split normalization into batches
ceiling(ncol(d_norm[["Data"]]) / 100)
dl <- list(
  "b1" = list(
    d_norm[["Data"]],
    d_norm[["Feature"]],
    d_norm[["Pixel"]]
  )
)
ncol(dl[["b1"]][[2]])
# Normalize data
# Repeat for each batch
d_norm2_b1 <- msi_data_norm(
  dm = dl[["b1"]][[1]],
  feat = dl[["b1"]][[2]],
  pix = dl[["b1"]][[3]],
  mtd = "sLOESS",
  parl = FALSE,
  core_perc = 0.9
)
# Recombine batches or save data
names(d_norm2_b1[["pixels"]])
ncol(d_norm2_b1[["data"]])
d_norm2 <- list(
  "data" = d_norm2_b1[["data"]],
  "features" = d_norm2_b1[["features"]],
  "pixels" = d_norm2_b1[["pixels"]][, -c(8)],
  "norm.method" = "sLOESS"
)
## Recalculate TIC for QC
d_norm2[["pixels"]][["TIC.norm.sloess"]] <- unlist(lapply(
  seq.int(1, ncol(t(as.matrix(d_norm2[["data"]]))), 1),
  function(x) sum(d_norm2[["data"]][x, ])
))
#---- Plot TIC ----
# view TIC
## Use lapply if on Windows
ptic <- msi_plot_tic(
  df = d_norm2[["pixels"]],
  var_x = "pixel",
  var_y = "TIC.norm.sloess",
  var_g = "ID",
  y_lim = 20000
)
ptic

## Plot ion images
ptic_img <- msi_plot_img(
  ptype = "intensity",
  df = d_norm2[["pixels"]],
  var_y = "TIC.norm.sloess",
  var_g = "ID",
  perc_int = 0.99
)
ptic_img
```

# 3.1. Image Segmentation, Clustering, and Region Assignment
#### Implements the scRNA-Seq clustering approach used by the R package Seurat to perform segmentation analysis and clustering of MSI data. Briefly, Seurat utilizes a K-nearest neighbor graph to partition all data points into individual 'communities' and evaluates overlapping communities based on their Jaccard similarity scores. This algorithm uses annotated compounds as input and therefore reduces bias introduced by technical artifacts and dramatically reduces the time necessary to generate segmented images. Furthermore, this method reliably partitions regions in tissue types with highly heterogenous morphology (e.g. lung tissue).
```{r, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
#---- Image Segmentation ----
# Segment based on all features after artifact removal
# Either perform segmentation for all samples or individually
## For all samples
d_norm2[["data"]]
## For each sample (use lapply if on Windows)
l_seg <- setNames(
  lapply(
    seq.int(1, length(unique(as.character(d_norm2[["pixels"]][["ID"]]))), 1),
    function(x) {
      s1 <- msi_segment(
        df = t(
          as.matrix(d_norm2[["data"]][d_norm2[["pixels"]][["ID"]] == x, ])
        ),
        md = d_norm2[["pixels"]][d_norm2[["pixels"]][["ID"]] == x, ],
        var_id = "ID",
        clus_res = 0.9
      )
      return(s1)
    }
  ),
  paste(
    rep("sample.",
      length(unique(as.character(d_norm2[["pixels"]][["ID"]])))
    ),
    unique(as.character(d_norm2[["pixels"]][["ID"]])),
    sep = ""
  )
)
## Visualize Clusters
msi_umap_panel(
  l_seg[[2]],
  c("ID", "Group", "cluster"),
  "umap"
)

#---- Find marker compounds for each cluster ----
p_hmap_top <- msi_marker_heatmap(
  so = l_seg[[2]],
  h_w = 32,
  h_h = 16,
  fs_c = 6,
  fs_r = 8,
  cl_c = TRUE,
  cl_r = TRUE,
  rot_c = 45,
  col1 = col_grad()[c(3, 6, 9, 12)]
)
p_hmap_top[["markers"]]
p_hmap_top[["plot"]]

#---- Plot clusters ----
# Add Seurat clustering column to image data frame for each sample and overwrite
col_reg <- data.frame(
  "ID" = unlist(lapply(
    seq.int(1, length(l_seg), 1),
    function(x) l_seg[[x]]@meta.data[["ID"]]
  )),
  "pixel" = unlist(lapply(
    seq.int(1, length(l_seg), 1),
    function(x) l_seg[[x]]@meta.data[["pixel"]]
  )),
  "cluster" = unlist(lapply(
    seq.int(1, length(l_seg), 1),
    function(x) l_seg[[x]]@meta.data[["cluster"]]
  ))
)
d_seg <- list(
  "data" = d_norm2[["data"]],
  "features" = d_norm2[["features"]],
  "pixels" = dplyr::left_join(
    d_norm2[["pixels"]],
    col_reg[, c("pixel", "cluster")],
    by = "pixel"
  ),
  "norm.method" = "sLOESS"
)

# Plot clusters
## (either all samples, individual sample, or all clusters for one sample)
names(d_seg[["pixels"]])
p_reg_img <- msi_plot_img(
  ptype = "regions",
  df = d_seg[["pixels"]],
  var_y = "cluster",
  var_g = "ID",
  perc_int = 0.95,
  spl_samp = FALSE,
  spl_reg = FALSE,
  samp_no = 1
)
p_reg_img

# Add annotated regions to data
d_seg[["pixels"]][["region"]] <- ifelse(
  d_seg[["pixels"]][["ID"]] == 1 & d_seg[["pixels"]][["cluster"]] == 1 |
    d_seg[["pixels"]][["ID"]] == 2 & d_seg[["pixels"]][["cluster"]] == 1,
  "region1",
  d_seg[["pixels"]][["cluster"]]
)
unique(d_seg[["pixels"]][, c("ID", "region")])
```

# 3.2. Colocalization to specific region
#### Plots ion images with cluster overlays to determine compound localization to a specific region.
```{r, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
# Plot ion images with region overlay
d_seg[["data"]] <- setNames(
  as.data.frame(d_seg[["data"]]),
  paste("X", seq.int(1, ncol(d_seg[["data"]]), 1), sep = ".")
)
d_seg[["data"]]
coloc1 <- msi_plot_img(
  ptype = "coloc",
  df = d_seg[["data"]],
  dfmd = d_seg[["pixels"]],
  var_y = "X.6",
  var_g = d_seg[["pixels"]][["ID"]],
  perc_int = 0.95,
  reg_var = "region",
  reg1 = "region1"
)
coloc1
```

# 4.1. Statistical Analysis
#### Statistical methods for comparing compound intensities between samples, regions, or a combination of variables.
```{r, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
# Count pixels per group
cnt_pix <- msi_cnt_pix(
  mdf = d_seg[["pixels"]],
  var_sel = c("Group", "ID", "cluster")
)
head(cnt_pix)

# Log-transform and scale data for statistical analysis
d_seg[["scale.data"]] <- msi_data_check()
head(d_seg[["scale.data"]][["data"]][["data.pareto"]])
# Conduct either one-way or multivariate ANOVA for
# segmented MSI dataset
## Subset/aggregate data beforehand if comparing individual regions
## within samples
head(d_seg[["pixels"]])
d_stat1 <- msi_stat_anova(md_var = c("Group"))
head(d_stat1)

## Repeat with aggregated pixels (compare treatment group means)
## Add or remove variables as needed
## use lapply if using Windows
### IMPORTANT: Must have at least 3 samples/group to work correctly
d_seg2 <- list(
  "data.pareto" = purrr::reduce(
    lapply(
      seq.int(1, ncol(d_seg[["data"]])),
      function(x) {
        setNames(aggregate(
          d_seg[["scale.data"]][["data"]][["data.pareto"]][[x]],
          list(d_seg[["pixels"]][, "ID"]),
          function(y) mean(y)
        ), c("ID", d_seg[["features"]][x, "Name"]))
      }
    ),
    dplyr::left_join,
    by = "ID"
  )[, -1],
  "data" = purrr::reduce(
    lapply(
      seq.int(1, ncol(d_seg[["data"]])),
      function(x) {
        setNames(aggregate(
          d_seg[["data"]][[x]],
          list(d_seg[["pixels"]][, "ID"]),
          function(y) mean(y)
        ), c("ID", d_seg[["features"]][x, "Name"]))
      }
    ),
    dplyr::left_join,
    by = "ID"
  )[, -1],
  "features" = d_seg[["features"]],
  "pixels" = unique(d_seg[["pixels"]][, c("Group", "ID")]),
  "norm.method" = "sLOESS"
)
d_stat <- msi_stat_anova(
  matpv = d_seg2[["data.pareto"]],
  matfc = d_seg2[["data"]],
  md = d_seg2[["pixels"]],
  md_var = c("Group"),
  an = d_seg2[["features"]]
)
head(d_stat)

# Plot statistical results as a volcano plot
p_vol <- msi_plot_volcano(
  d_stat = d_stat1,
  comp_name = "M.SFA-M.HO3",
  f_cut = 0.25,
  f_lim = 6,
  y_limit = 10,
  x_title = "x-axis title"
)
p_vol
```

# 4.2. ChemRICH
#### Conducts ChemRICH analysis for each comparison in a provided result generated by msi_stat_anova() (default) or from a single ChemRICH input (in development).
```{r, echo=T, fig.align= "center", results=T, message=T, warning=T, eval=T}
# Run ChemRICH
## IMPORTANT: only works for datasets with annotated compounds
d_seg2[["features"]][["label.test"]] <- ifelse(
  d_seg2[["features"]][["Count"]] < 100,
  "low",
  "high"
)
d_enr <- msi_stat_crich(
  d_stat = d_stat1,
  d_ref = d_seg2[["features"]],
  cl_name = "label.test"
)
head(d_enr)
# Plot
unique(d_enr[["Comparison"]])
p_enr <- msi_plot_crich(
  dfcr = d_enr[d_enr[["Comparison"]] == "M.SFA-M.HO3", ],
  cl_name = "label.test"
)
p_enr
```
